---
output: pdf_document
---

# SNA with R-igraph: a tutorial with enron e-mails dataset

Written by [Pedro Concejero](https://www.linkedin.com/in/pedroconcejero) based on previous work by  [Carlos Ortega](http://www.qualityexcellence.es/) for [Madrid R users group](http://madrid.r-es.org/). This session was hel in Madrid on April 2013, was in spanish and is recorded so you can watch it [here](https://www.intecca.uned.es/portalavip/grabacion.php?ID_Sala=77088&ID_Grabacion=80882&hashData=497fb4164f5080c3a94124de570f6707&amp%3BparamsToCheck=SURfR3JhYmFjaW9uLElEX1NhbGEs)

This document reviews the original 2013 tutorial material, but with appropiate updates (ie links) on august 2015. All code and datasets are available at my [github](https://github.com/pedroconcejero/public-sna).

(check from root https://github.com/pedroconcejero in case you fail to find datasets because I must change the structure of github in the near future).

You can contact me here: pedro.concejerocerezo *at* gmail.com. I will be delighted to listen to your comments!!!

This document is done in RStudio using [knitr markdown](http://rmarkdown.rstudio.com/) language and intends to provide a tutorial about the igraph R library for Social Networks Analysis.

Explanations on how to use igraph for producing R graph objects are embedded within the explanations of the data and objectives of the analysis.

# Introduction: The enron scandal

The enron skandal revealed in 2001 and was the most expensive bankruptcy produced till that date (many more expensive ones have happened afterwards). An excellent reference on the enron history can be found in wikipedia:

[http://en.wikipedia.org/wiki/Enron_scandal](http://en.wikipedia.org/wiki/Enron_scandal)

After the company's collapse a large database of over 600,000 emails generated by 158 employees of the Enron Corporation was acquired by the Federal Energy Regulatory Commission during its investigation after the company's collapse. A copy of the database was subsequently purchased for $10,000 by Andrew McCallum, a computer scientist at the University of Massachusetts Amherst, who released this copy to researchers as the "Enron corpus". This analysis is based on this dataset. More about the Enron corpus can be consulted at:

[http://en.wikipedia.org/wiki/Enron_Corpus](http://en.wikipedia.org/wiki/Enron_Corpus)

*******

# The enron e-mail database 

The dataset object of analysis here is based on a MySQL implementation of all e-mails between the 158 enron employees and all the rest of the world (except private emails that were deleted previously by the database owners). Since this dataset would be rather difficult to use in an educational setting, the dataset was restricted to the emails **between enron employees**, i.e. both sender and receiver are enron employees, thus reducing considerably the dataset size and making it easier the interpretation of links and other practical issues.

This dataset was created from a version of the enron corpus by Jitesh Shetty and Jafar Adibi available as a MySQL database [still available here](http://www.cs.cmu.edu/~./enron/). Please note this database was reviewed in several ocasions after the original tutorial was produced, and authors recommend to download and use the latest version. More in particular, these [deletions](http://www.cs.cmu.edu/~./enron/DELETIONS.txt). These deletions *have not been done* in this august 2015 version but I will try to implement them.

This is the origin of the two dataframes that are required to produce an igraph R graph object: edges (or links, in this case, e-mails), and nodes.

# Setting up the R + igraph environment

Needless to say (or maybe not), you need the igraph library installed in your system. If you have not do so, it is easy, once you have opened R, just introduce this sentence:

install.packages("igraph")

First of all we must prepare the R environment, load igraph and other required libraries, and set the working directory. Please change this for your environment!!!

```{r}

library(igraph)
# required to improve plots with function colorpanel
library(gplots)

setwd("D:/SNA_data_and_literature/enron")

```

You just need two files that contain all data required for these analyses:
- [enron.RData file](https://github.com/pedroconcejero/public-sna/blob/master/enron.RData) contains the working space with several R objects that are explained and described below
- [edges_w_message.RData (please unzip data first!)](https://github.com/pedroconcejero/public-sna/blob/master/edges_w_message.zip) is a file that includes a dataframe containing the links. 

For igraph it is essential that two first columns found in the edges dataframe are node id's -usually first one is sender and second is receiver-. So we have the following information in the edges dataframe:
- sender: e-mail address of sender
- receiver: e-mail address of receiver
- type of e-mail (CC, BCC, TO)
- subject: string with the subject of e-mail
- body: full text of e-mail message
- date

Date format in particular is tricky, you will find some cautions about it in the tutorial.

```{r results ='hide'}

load("enron.RData")
load("edges_w_message.RData")

```

We can do a few basic manipulations with these objects:

- Count the number of edges:

```{r}
nrow(edges.full)

```

- Describe the edges object with str function:

```{r}
str(edges.full)

```

- or we can re-format date string so that we can use dates (ie filter, compute durations) more easily in R.
```{r}
edges.full$date.R <- as.POSIXct(edges.full$date)
summary(edges.full$date.R)

```

Note that date can be either a string or a native date R-object. We will use both, because external software, in particular gephi, does not understand the exported R-date format.

The other required object to produce an igraph graph object is the nodes dataframe. This contains all the info about the nodes, in our case, the enron employees who were e-mail sender or receivers.

This dataframe contains e-mail address as node id, the lastName as a useful string for labelling, and her/his status in the company (if this info. was available).

```{r}
# Number of nodes
nrow(nodes)

# Description of the nodes object
str(nodes)
```

The rest of this document explains how to handle and what you can do with the igraph SNA object in 8 steps

**********
*********
# R igraph tutorial in 8 steps

## 1- Creating an igraph network graph with graph.data.frame

Just insist on the requirement that:
- two first columns of edges object match with node id's
- nodes object must contain all info. from nodes in edges object

When creating the graph we can choose if the network is directed or not. In this case we choose it as directed.

Please note that we create an igraph object and summary will tell us the number of  nodes and edges.
igraph automatically sets as node properties all additional columns in node object (name, lastName, status) and as edge properties all additional columns apart from node id's (type, date, count)

And note that for igraph V = vertex and E = edge. Please note uppercase.
And we filtered out the full text for practical reasons, to make it simpler.

```{r}

network.full <- graph.data.frame(edges.full[,c("sender",
                                               "receiver",
                                               "type",
                                               "date",
                                               "subject")],
                            directed = TRUE,
                            vertices = nodes)

class(network.full);summary(network.full)


```

********

## 2- Using the igraph object and V and E components

Best documentation can be found at:

 [http://igraph.sourceforge.net/doc/R/00Index.html](http://igraph.sourceforge.net/doc/R/00Index.html)
 
 [http://igraph.sourceforge.net/documentation.html](http://igraph.sourceforge.net/documentation.html)

And also from the unfinished tutorial:

 [http://igraph.sourceforge.net/igraphbook/](http://igraph.sourceforge.net/igraphbook/)

You can access node and edge properties by means of:

 V(your_network_object) 
 
 E(your_network_object)

```{r}
V(network.full)[1:10]
E(network.full)[1:10]

```

And also its properties via table

```{r}
table(V(network.full)$status)
```

********

## 3- Exporting the graph for using it with external software (ie gephi)

Basically we write the graph to graphml format (an XML format special for graph objects). Take care with date format: gephi requires it to be a string.

Gephi tutorial is not part of this document. You can find excellent gephi tutorials just [googling](https://www.google.es/search?q=gephi+tutorial)


```{r}

write.graph(network,
            file = "network01.graphml",
            format = "graphml")

```

******

## 4- Computing Individual SNA metrics

With igraph and get.shortest.paths you can obtain the **shortest paths** between two nodes.

Thanks to explanation at:

[http://sigloxxi.fcie.uam.es/informatica/media/Grafos%20con%20R%20e%20Igraph.pdf](http://sigloxxi.fcie.uam.es/informatica/media/Grafos%20con%20R%20e%20Igraph.pdf)

```{r}
get.shortest.paths(from = V(network.full)$lastName == "Pereira",
                   to = V(network.full)$lastName == "Horton",
                   graph = network.full)

nodes[c(138,11,132),]
```

**Diameter** of the graph is the length of the largest distance between nodes

```{r}

diameter(network.full)
nodes[farthest.nodes(network.full),]

```

**Centrality measures** are computed and can be added to the node properties table.
Basic centrality measure is **degree**, both in_degree and out_degree (this is a directed graph), and total_degree.

```{r}

nodes$degree_total <- degree(network.full, 
                             v = V(network.full), 
                             mode = c("total"))
nodes$degree_in <- degree(network.full, 
                             v = V(network.full), 
                             mode = c("in"))
nodes$degree_out <- degree(network.full, 
                             v = V(network.full), 
                             mode = c("out"))
```

Let's see who are the top20 for each measure

```{r}
head(nodes[order(nodes$degree_total,
           decreasing = TRUE),], n = 20L)

head(nodes[order(nodes$degree_in,
                 decreasing = TRUE),], n = 20L)

head(nodes[order(nodes$degree_out,
                 decreasing = TRUE),], n = 20L)

```

**Reach** is another measure, also known as **neighborhood.size**. You must specify a specific order (an integer), meaning the total number of people you can reach with that number of steps. We can observe how this metric is very much linked to actual connectivity.

```{r}
nodes$reach_2_step <- 
  neighborhood.size(network.full, 
                    order = 2,
                    nodes = V(network.full), 
                    mode  = c("all"))

head(nodes[order(nodes$reach_2_step,
                 decreasing = TRUE),], n = 30L)

```

There is a lot of info. about enron employees that can guide you for further exploring the enron social network, ie 
[http://www.inf.ed.ac.uk/teaching/courses/tts/assessed/roles.txt](http://www.inf.ed.ac.uk/teaching/courses/tts/assessed/roles.txt)

Other interesting measures are **clustering coefficient** and **transitivity** [wikipedia on transitivity measures](http://en.wikipedia.org/wiki/Clustering_coefficient). For instance: "The clustering coefficient places more weight on the low degree nodes, while the transitivity ratio places more weight on the high degree nodes".

```{r}

nodes$transitivity_ratio <- 
  transitivity(network.full, 
               vids = V(network.full), 
               type = "local")

head(nodes[order(nodes$transitivity_ratio,
                 decreasing = FALSE),], n = 20L)

V(network.full)$outdegree <- degree(network.full, mode = "out")
V(network.full)$indegree <- degree(network.full, mode = "in")
V(network.full)$degree <- degree(network.full, mode = "all")
V(network.full)$reach_2_step <-   neighborhood.size(network.full, 
                                                    order = 2,
                                                    nodes = V(network.full), 
                                                    mode  = c("all"))
V(network.full)$transitivity_ratio <- transitivity(network.full, 
               vids = V(network.full), 
               type = "local")

```

********

## 5- Extracting subgraphs

Extracting parts of a graph using igraph is very easy. You just need to know two functions:

induced.subgraph

subgraph.edges

For instance, to extract subgraphs of the most relevant people when enron came into bankruptcy. Many details about this [here](http://es.wikipedia.org/wiki/Enron#Ca.C3.ADda_de_la_empresa) (CAVEAT: in spanish, as I have not been able to find this information with such detail in english)


```{r}

edges.full$day <- strftime(edges.full$date.R, "%Y-%m-%d")

network.august <- subgraph.edges(network.full,
                                 which(as.Date(E(network.full)$date) > "2001-02-12 00:00:00"),
                                 delete.vertices = TRUE)
```

We actually produce a new graph 

```{r}

summary(network.august)
```

For instance let's see messages from president Kenneth Lay

```{r}
mails.lay <- edges.full[(edges.full$sender == "kenneth.lay@enron.com" &
                           as.Date(edges.full$date.R) > "2001-07-01 00:00:00") |
                          (edges.full$receiver == "kenneth.lay@enron.com" &
                             as.Date(edges.full$date.R) > "2001-07-01 00:00:00") 
                        ,]
mails.lay <- mails.lay[order(as.Date(mails.lay$date.R)),]
nrow(mails.lay)

```

See how employees were not aware until last minute of what was going on, in spite of all the stakes they had in the company performance. But of course all this depended on the position you had in the company.

This is the email to Kenneth Lay from a young employee who lost money after believing in enron executives communications (though they were doing the opposite):

```{r}
mails.lay[rownames(mails.lay) == 3473,]

```

And this is the message of an executive to the same. Tone or words cannot be more different.

```{r}
mails.lay[rownames(mails.lay) == 60469,]

```

Another way of extracting a subgraph, all nodes who had contact with Kenneth Lay 

```{r}
nodes.with.lay <- unique(c(mails.lay$sender,
                           mails.lay$receiver))

network.kenneth.lay <- graph.data.frame(mails.lay[,c("sender",
                                          "receiver",
                                          "type",
                                          "date",
                                          "subject")],
                                        directed = TRUE)

summary(network.kenneth.lay)

```

And now see how many people were in Lay's neighbourhood. This was the CEO so it was extremely easy for him to reach the whole company in only two steps.

```{r}
neighborhood.size(network.full, 
                  1, 
                  V(network.full)$lastName == "Lay")


neighborhood.size(network.full, 
                  2, 
                  V(network.full)$lastName == "Lay")

```


****

## 6- Reciprocity and Dyads - Creating the social graph 

The [reciprocity](http://igraph.sourceforge.net/doc-0.5.1/R/reciprocity.html) function gives this measure.

```{r}
reciprocity(network.full)

```

And you can also obtain the [dyads](http://igraph.sourceforge.net/doc-0.5.1/R/dyad.census.html)

This gives back A named numeric vector with three elements:

 mut   The number of pairs with mutual connections.
 
 asym  The number of pairs with non-mutual connections.
 
 null   The number of pairs with no connection between them.

Similar thing can be done with triplets (not done here). See [http://igraph.sourceforge.net/doc-0.5.1/R/triad.census.html](http://igraph.sourceforge.net/doc-0.5.1/R/triad.census.html)
 
```{r}
dyad.census(network.full)

```

Social graph should contain reciprocal pairs. This is, our directed pairs A->B would require that we also have a relationship B->A. This way we would have a reciprocal relationship A<->B.
This way we convert a **communications graph** into a **social graph**

For that purpose we need a preliminary step: computing the weight of the link between two nodes. The simplest measure is the number of communications without distinction by type (to, cc, bcc).

First we extract unique pairs and we order them:

```{r}
pairs <- as.data.frame(unique(edges.full[c(1,2)]))
pairs <- pairs[order(pairs$sender, pairs$receiver),]

edges.ordered <- edges.full[order(edges.full$sender, edges.full$receiver),]

weight <- aggregate(edges.ordered[,3],
                by = list(edges.ordered[,1],
                          edges.ordered[,2]),
                length) 

weight <- weight[order(weight$Group.1, weight$Group.2),]

```

Let's verify with head and tail 

```{r}
head(pairs, n = 5L)
head(weight, n = 5L)
tail(pairs, n = 5L)
tail(weight, n = 5L)
pairs[seq(236:248),]
weight[seq(236:248),]

```

Then we mix pairs and weight 

```{r}
pairs$weight <- weight$x
head(pairs)
```

Now we substitute the mails table by a links table and we produce a new graph using this as the edge table

```{r}
network.sna <- graph.data.frame(pairs,
                            directed = TRUE,
                            vertices = nodes)

summary(network.sna)

```

Let's check with reciprocity and dyad.census

```{r}
reciprocity(network.sna)
dyad.census(network.sna)
```

Now we can impose the requirement that a link must be reciprocal to have a **social relationship**

Thanks to [Carlos Gil Bellosta](http://www.datanalytics.com/) for suggesting:

[http://stackoverflow.com/questions/13006656/igraph-nonreciprocal-edges-after-converting-to-undirected-graph-using-mutual](http://stackoverflow.com/questions/13006656/igraph-nonreciprocal-edges-after-converting-to-undirected-graph-using-mutual)

[http://igraph.sourceforge.net/doc/R/as.directed.html](http://igraph.sourceforge.net/doc/R/as.directed.html)

```{r}
network.social <- as.undirected(network.sna, 
                                mode = "collapse", 
                                edge.attr.comb = "sum")
```



*****

## 7- Computing Communities

You should know there are recommendations about availability of [algorithms for computing communities](http://igraph.wikidot.com/community-detection-in-r) depending on the type of your graph (directed vs. non-directed).

A classical algorithm is the one by Vincent D Blondel, Jean-Loup Guillaume, Renaud Lambiotte, Etienne Lefebvre, "Fast unfolding of communities in large networks", in Journal of Statistical Mechanics: Theory and Experiment 2008 (10), P1000, and is part of gephi in the function [multilevel.community](http://igraph.sourceforge.net/doc/R/multilevel.community.html)

```{r}
communities <- multilevel.community(network.social)

comms.df <- data.frame(row.names = seq(1:149))
comms.df$Email_id <- communities$names
comms.df$community <- communities$membership

```

Then you can add each node's community to the nodes table

```{r}
nodes.def <- merge(nodes, comms.df, 
                   by.x = "Email_id",
                   by.y = "Email_id")

str(nodes.def)
head(nodes.def)
barplot(table(nodes.def$community))

V(network.social)$community <- communities$membership

```


******

## 8- Graph visualization (plotting) with igraph

Let's try plot quick and dirty:

```{r fig.width=10, fig.height=10}
plot(network.social)

```

First recommendation: 

**plotting a large graph with igraph -and the enron graph is relatively small- is useless** 

Gephi is an excellent alternative for high quality plots of relatively large graphs, but anyway the smaller, more compact and especially more relevant the better.

So the first thing we shall do is extracting a relevant subgraph. Let's extract the "CEO's COMMUNITIES":

```{r fig.width=10, fig.height=10}
str(nodes.def)
nodes.def[nodes.def$lastName == "Lay",]

# uncomment if you want to see all details of community number 8
#nodes.def[nodes.def$community == 8, c(2:9)]

com.ceos <- induced.subgraph(network.social,
                             V(network.social)$community == 8,
                             impl = "auto") # Ver ayuda

summary(com.ceos)


g <- com.ceos

plot(g,
     vertex.label.font = 1,
     vertex.label.cex = 0.8, 
     edge.arrow.size = 0.3,
     vertex.size = 12,
     vertex.color = "yellow"
)

```

You can see another particularity: By default (no layout), nodes are projected on *random co-ordinates*, with automatic labels starting by 0, correlative numbers afterwards. Therefore every time you plot the graph, the result will be different. If you want to keep stable a plot you must fix the layout, or you can use some pre-computed layouts with fancy names, as "Reingold-Tilford" (after its authors), which tell you nothing about what they do (so you just have to try and see) or some others who are more clearly labelled, as the layout.circle. 

Let's see some of them. Please note this was prepared in 2013 and layouts have changed in igraph. Always check latest documentation to see this! I have updated the original tutorial with this info, but it can still change.

### How to fix a layout in a plot

First the definition of layout:
Layout - Either a function or a numeric matrix. It specifies how the vertices will be placed on the plot.
If it is a numeric matrix, then the matrix has to have one line for each vertex,  specifying its coordinates. The matrix should have at least two columns, for the x and y coordinates, and it can also have third column, this will be the z coordinate for 3D plots and it is ignored for 2D plots.

If a two column matrix is given for the 3D plotting function rglplot then the third column is assumed to be 1 for each vertex.

If layout is a function, this function will be called with the graph as the single parameter  to determine the actual coordinates. The function should return a matrix with two or three columns. 

For the 2D plots the third column is ignored.

Let's fix a random layout. Note: the layout is still random, what you do with this code is to produce the same result every time you use the plot function.

```{r fig.width=12, fig.height=12}
l <- layout.random(g)
plot(g,layout = l)
```

A preliminary conclusion: even for small graphs a random layout might not be the best representation.

Let's check some other layouts.

### Reingold-Tilford layout

Reingold.tilford produces a hierarchical graph. Note! This function is deprecated, it might issue a warning.

```{r fig.width=12, fig.height=12}
plot(g, 
     main = "enron graph with hierarchical layout", 
     layout = layout.reingold.tilford,
     vertex.label = V(g)$label)
```


### Circle layout (Chord plot)

layout.circle produces circle or also known as chord plots. Note! This function is deprecated, it might issue a warning.

```{r fig.width=12, fig.height=12}
l <- layout.circle(g)

# use colour functions
par(bg = "#000000", mar = c(1,1,1,1), oma= c(1,1,1,1))

edge_col <- colorpanel(length(table(E(g)$weight)), 
                       low = "#2C7BB6", high = "#FFFFBF")  
E(g)$color <- edge_col[factor(E(g)$weight)]

plot(g, 
     layout=l,
     vertex.label = V(g)$label,
     vertex.size=1,
     vertex.label.color = "white",
     edge.width = 1.5*log10(E(g)$weight), 
     edge.curved = F, 
     edge.color = E(g)$color)

```

### Other layouts: fruchterman.reingold and kamada.kawai


```{r fig.width=12, fig.height=12}
# Let't try to improve the position of objets in plain
# Let's put name as the node label

V(g)$label <- V(g)$lastName

# This is fruchterman reingold

plot(g, 
     layout = layout.fruchterman.reingold, 
     vertex.label.font = 1,
     vertex.label.cex = 0.8, 
     edge.arrow.size = 0.3,
     vertex.size = 12,
     vertex.color = "yellow")

# This is kamada.kawai

plot(g,
     layout=layout.kamada.kawai)

```

Let's improve these plots

```{r fig.width=12, fig.height=12}

# color the edges

par(bg = "#000000", mar = c(1,1,1,1), oma= c(1,1,1,1))

edge_col <- colorpanel(length(table(E(g)$weight)), 
                       low = "#2C7BB6", high = "#FFFFBF")  
E(g)$color <- edge_col[factor(E(g)$weight)]


plot(g, 
     main = "enron", 
     layout = layout.fruchterman.reingold(g, 
                                          params= list(niter = 1000, 
                                                       weights = E(g)$weight)),
     vertex.label = V(g)$label, 
     vertex.size = log10(as.numeric(V(g)$degree_total)),
     vertex.label.font = 1,
     vertex.label.color = "white",
     vertex.label.cex = 0.8, 
     edge.arrow.size = 0.3,
     vertex.color = "yellow",
     edge.arrow.size = E(g)$weight/150, 
     edge.width = 1.5*log10(E(g)$weight), 
     edge.curved = T, 
     edge.color = E(g)$color)

plot(g, 
     main = "enron", 
     layout = layout.kamada.kawai(g, 
                                          params= list(niter = 1000, 
                                                       weights = E(g)$weight)),
     vertex.label = V(g)$label, 
     vertex.size = 12,
     vertex.label.font = 1,
     vertex.label.color = "black",
     vertex.label.cex = 0.8, 
     edge.arrow.size = 0.3,
     vertex.color = "yellow",
     edge.arrow.size = E(g)$weight/150, 
     edge.width = 1.5*log10(E(g)$weight), 
     edge.curved = T, 
     edge.color = E(g)$color)



```

